Wildcard: Modifying web applications by directly manipulating a data table

### 2019-12-28 paper.tex
done for the day

\title{{[}DRAFT, please don't distribute{]} Wildcard: a tool for
modifying websites by directly manipulating data tables}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Browser extensions and user scripts can modify websites in useful
ways---ranging from blocking ads to adding entire new features to
Gmail---but many people have unique needs that aren't met by existing
extensions. Today, most of those people are stuck. They can't build
their own browser extensions without learning how to program, so they
have no choice but to accept the way the software was built. What if
things were different?

Wildcard is a platform that empowers anyone to build browser extensions
and modify websites to meet their own specific needs. Wildcard shows a
simplified view of the data in a web page as a familiar table view.
People can directly manipulate the table to sort/filter content, add
annotations, and even use spreadsheet-style formulas to pull in data
from other websites. The key idea is that a table is a powerful, simple,
and familiar paradigm for modifying a website.
\end{abstract}

@@

Todos in intro:

Do somes tuff

\begin{itemize}
\tightlist
\item
  hook with Airbnb sort
\item
  add more reasons
\item
  make this shorter
\end{itemize}

@@

There have been many research projects that try to address these
hurdles. For example, Chickenfoot \autocite{miller2010} allows for
people to more quickly modify sites without using Javascript or dealing
with the DOM, directly addressing both of these barriers. But, a decade
later, not many people have ended up using systems like this.

\hypertarget{why-dont-people-tweak-websites}{%
\subsubsection{Why don't people tweak
websites?}\label{why-dont-people-tweak-websites}}

@@

There's somewhat of a tension here between directness and structure.
Manipulating the original page itself might seem most ``direct.'' But
the whole problem we're dealing with is that the original web page
doesn't provide affordances for end-user modification, and there's no
consistent structure that people can learn to work with across many
sites.

\hypertarget{directness}{%
\subsubsection{Directness}\label{directness}}

The table view is perhaps one hop less ``direct,'' but in turn provides
other benefits. Because many people are already familiar with
spreadsheets, they can quickly intuit which changes are easy to make in
this system. Also, people can learn to work with the same consistent
spreadsheet view across many sites. This consistency is very
important---learning a generic tool that can be applied to many
different specific cases is very powerful, and connected to the idea of
literacy in a medium (just ask anyone who has used vim, or spreadsheets,
or a pencil and paper). Wildcard aims to balance this tension, with a
workflow that involves both the table view and the original webpage.

@@

\hypertarget{the-big-picture}{%
\subsubsection{The big picture}\label{the-big-picture}}

@@

Again, ultimately we hope that first-party sites would find it
beneficial and straightforward to provide a structured data view
themselves. Wildcard doesn't require sites to expose some complex
Semantic Web schema; it merely asks for a simple structured data view.

\hypertarget{demo}{%
\section{Demo}\label{demo}}

@@

\hypertarget{expose-generic-structure}{%
\subsection{Expose generic structure}\label{expose-generic-structure}}

\begin{itemize}
\tightlist
\item
  Consistent across all sites, can gain familiarity
\item
  Nightmare bike: users learn the structure they can exploit
\end{itemize}

\hypertarget{no-api-needed}{%
\subsection{No API needed}\label{no-api-needed}}

\begin{itemize}
\tightlist
\item
  3p-only is enough. 1p help is optional.
\item
  Can even do things the 1p didn't want to expose.
\end{itemize}

\hypertarget{in-place-toolchain}{%
\subsection{In-place toolchain}\label{in-place-toolchain}}

\begin{itemize}
\tightlist
\item
  meet the user where they are, in the page
\item
  very native to the web style (dev tools is kinda like this)
\end{itemize}

@@

\hypertarget{analysis}{%
\section{Analysis}\label{analysis}}

Cognitive dimensions?

@@

\hypertarget{future-work-open-questions}{%
\section{Future work / Open
Questions}\label{future-work-open-questions}}

@@

  \begin{itemize}
  \tightlist
  \item
    workflows? triggers?
  \item
    what can and can't be done?

    \begin{itemize}
    \tightlist
    \item
    \end{itemize}
  \end{itemize}
\item
  automated wrappers?

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

\hypertarget{todos}{%
\section{Todos}\label{todos}}

\begin{itemize}
\tightlist
\item
  remove bold text for the proceedings version
\end{itemize}

>>>>

\title{Wildcard: Modifying web applications by directly manipulating a
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Browser extensions and user scripts can modify web applications in
useful ways, but many people have unique needs that aren't met by
existing extensions. Today, most of them are stuck---they can't build
their own browser extensions without learning how to program in
Javascript.

Wildcard is a platform that empowers anyone to build browser extensions
without doing traditional programming. Wildcard shows the main data from
a web page in a table with a bidirectional connection to the original
page. People can directly manipulate the table to sort/filter content,
add private annotations, and use spreadsheet formulas to fetch data from
other websites.
\end{abstract}

@@

Todos for the introduction:

\begin{itemize}
\tightlist
\item
  hook with Airbnb sort
\item
  add more bigger picture reasons, less focus on the affordances
\item
  make this shorter
\end{itemize}

@@

There have been many research projects that try to address these
hurdles. For example, Chickenfoot \autocite{bolin2005} allows for people
to more quickly modify sites without using Javascript or dealing with
the DOM, directly addressing both of these barriers. But, a decade
later, not many people have ended up using systems like this.

@@

@@

\hypertarget{demo-booking-a-trip-with-wildcard}{%
\section{Demo: Booking a trip with
Wildcard}\label{demo-booking-a-trip-with-wildcard}}

\hypertarget{implementation}{%
\section{Implementation}\label{implementation}}

@@

@@

Here are some of the ideas behind the design of Wildcard. We hope these
can be useful principles not only for Wildcard, but also for other tools
that enable end users to modify software.

\hypertarget{expose-unified-structure-across-applications}{%
\subsection{Expose unified structure across
applications}\label{expose-unified-structure-across-applications}}

In \emph{Changing Minds} \autocite{disessa2000}, Andrea diSessa
critiques the design of modern software with a story about a
hypothetical ``nightmare bike.'' Each gear on the nightmare bike is
labeled not with a number, but with an icon describing its intended use:
smooth pavement uphill, smooth pavement downhill, gravel, etc. By some
logic, this is more ``user-friendly'' than numbered gears, but in fact,
hiding orderly structure from the user makes it more difficult to
operate the bike. Many modern software designs fall into this trap,
teaching users to use isolated modes rather than coherent structure, and
the problem gets far worse when operating across multiple applications.
Unlike the UNIX philosophy of small tools interoperating through shared
abstractions, in modern computing each application is in its own silo of
data and functionality.

Wildcard helps people understand and modify the behavior of applications
through the lens of a consistent abstraction: a data table. This
abstraction strikes a balance between being both simple and generic. A
data table is simpler than the DOM tree that describes the details of
the UI, but is also generic enough to describe the essence of many
different kinds of applications.

Exposing a unified higher-level abstraction is a deliberate choice, and
is not the only way to enable users to program without directly using
the DOM. Chickenfoot \autocite{bolin2005} and CoScripter
\autocite{leshed2008} allow users to create scripts that resemble
natural language and ``sloppily'' match queries to elements in the DOM.
These designs allow for a wide range of operations, but don't explicitly
indicate what operations are possible---the user must look at the
original page and imagine the possibilities. In contrast, Wildcard
provides affordances that clearly suggest the availability of certain
types of modifications. In addition to giving users more certainty about
whether a modification is possible, these affordances might give users
new ideas for things to try. Because people are not used to modifying
web applications, providing inspiration is an important goal. (Todo:
Perhaps something to cite here, re: discoverability in GUIs vs CLIs?)

\hypertarget{direct-manipulation-by-proxy}{%
\subsection{Direct manipulation by
proxy}\label{direct-manipulation-by-proxy}}

In Wildcard, users don't interact with the original page to modify it;
instead they manipulate an alternate representation of the data in the
page. The tool uses visual cues to indicate the mapping between the
table representation and the original page. This is similar to the way
that browser developer tools use in-page highlighting to indicate the
mapping between HTML and the original page.

\begin{figure}
\hypertarget{fig:devtools}{%
\centering
\includegraphics{media/devtools.png}
\caption{The Chrome Dev Tools use highlighting to show the mapping
between HTML code and the page}\label{fig:devtools}
}
\end{figure}

We considered alternate approaches where the user would interact more
directly with the original page, but decided that the table view
provided a consistent

\begin{itemize}
\tightlist
\item
  True DM might be right on the page itself

  \begin{itemize}
  \tightlist
  \item
    There are good UIs for this, eg Chrome DevTools element picker
  \end{itemize}
\item
  Considered for Wildcard, but went with an indirect proxy. Still
  maintain a close mapping.
\item
  Pros: consistency across sites. Avoid site-specific styling issues.
\item
  Cons: maybe less intuitive? More work to do the mapping, might be a
  source of confusion.
\end{itemize}

\hypertarget{first-party-optional}{%
\subsection{First party optional}\label{first-party-optional}}

\begin{itemize}
\tightlist
\item
  3p-only is enough. 1p help is optional.
\item
  Can even do things the 1p didn't want to expose.
\item
  make it easy for 1p: just ask for a data table, no fancy schema.
\end{itemize}

\hypertarget{built-for-the-web}{%
\subsection{Built for the web}\label{built-for-the-web}}

\begin{itemize}
\tightlist
\item
  web has the right foundations:

  \begin{itemize}
  \tightlist
  \item
    the right platform
  \item
    open platform
  \item
    dev tools
  \item
    originally intended to be
  \end{itemize}
\item
  in-place toolchain: meet the user where they are, in the page
\end{itemize}

@@

@@

\hypertarget{future-work}{%
\section{Future work}\label{future-work}}

@@

  \begin{itemize}
  \tightlist
  \item
    workflows? triggers?
  \item
    what can and can't be done
  \end{itemize}
\item
  automated wrappers?

@@

\hypertarget{meta-todos}{%
\section{meta: todos}\label{meta-todos}}

\begin{itemize}
\tightlist
\item
  remove bold text for the proceedings version
\item
  go through the Programming proceedings template checklist
\end{itemize}
